---
title: "HIV Study on the Entire African Continent"
author: "Uthman Al-Ameen Olalekan"
date: "2025-06-22"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Required Libraries
```{r include=FALSE}
library(tidyverse)
library(flextable)
library(officer)
library(scales)
library(stringi)
library(readr)
library(viridis)
library(ggpubr)
library(plm)
library(lme4)
library(car)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)
library(reactable)

```

## Load and View Raw Data
```{r}
africa <- read_csv("C:/Users/THIS PC/Downloads/africa.csv")
art_coverage_xlsx <- read_csv("C:/Users/THIS PC/Downloads/art_coverage.xlsx.csv")
aids_mortality_xlsx <- read_csv("C:/Users/THIS PC/Downloads/aids_mortality.xlsx.csv")
hiv_incidence_xlsx <- read_csv("C:/Users/THIS PC/Downloads/hiv_incidence.xlsx.csv")

```
## Filter Only African Countries
```{r}
african_countries <- africa %>%
  filter(`Location type` == "Country") %>%
  distinct(Location) %>%
  pull(Location)

# Add missing countries to make 54
missing_countries <- c("Egypt", "Libya", "Morocco", "Sudan", "Tunisia", "Somalia", "Djibouti")
african_countries <- c(african_countries, missing_countries)
african_countries <- data.frame(Country = african_countries)

```

## Clean and Filter Country Names in Datasets
```{r}
african_countries_clean <- african_countries$Country %>%
  stri_trans_general("Latin-ASCII") %>%
  tolower() %>%
  trimws()

filter_and_clean <- function(data, var = "Country") {
  data %>%
    mutate(Country_clean = stri_trans_general(!!sym(var), "Latin-ASCII") %>% tolower() %>% trimws()) %>%
    filter(Country_clean %in% african_countries_clean)
}

filtered_art_coverage <- filter_and_clean(art_coverage_xlsx)
filtered_hiv_incidence <- filter_and_clean(hiv_incidence_xlsx)
filtered_aids_mortality <- filter_and_clean(aids_mortality_xlsx)

```

## Save Filtered Datasets
```{r}
write_csv(filtered_art_coverage, "filtered_art_coverage.csv")
write_csv(filtered_hiv_incidence, "filtered_hiv_incidence.csv")
write_csv(filtered_aids_mortality, "filtered_aids_mortality.csv")

```

## Convert to Long Format
```{r}
art_long <- filtered_art_coverage %>%
  pivot_longer(cols = starts_with("20"), names_to = "Year", values_to = "ART_Coverage")

hiv_long <- filtered_hiv_incidence %>%
  pivot_longer(cols = starts_with("20"), names_to = "Year", values_to = "HIV_Incidence")

mortality_long <- filtered_aids_mortality %>%
  pivot_longer(cols = starts_with("20"), names_to = "Year", values_to = "AIDS_Mortality")

```

## Clean the Datasets
```{r}
clean_data <- function(data, value_col) {
  data %>%
    filter(grepl("^[0-9]{4}$", Year)) %>%
    select(Country, Year, !!sym(value_col)) %>%
    mutate(Year = as.integer(Year))
}

art_long_clean <- clean_data(art_long, "ART_Coverage")
hiv_long_clean <- clean_data(hiv_long, "HIV_Incidence")
mortality_long_clean <- clean_data(mortality_long, "AIDS_Mortality")

```

## Merge and Clean
```{r}
common_data <- art_long_clean %>%
  inner_join(hiv_long_clean, by = c("Country", "Year")) %>%
  inner_join(mortality_long_clean, by = c("Country", "Year"))

# Remove countries with missing key indicators
countries_to_remove <- c("Cameroon", "Central African Republic", "Equatorial Guinea",
                         "São Tomé and Príncipe", "Sao Tome and Principe", "Benin")

common_data_filtered <- common_data %>%
  filter(!Country %in% countries_to_remove)

```

## Adjust for "<" Scenario
```{r}
common_data_adjusted <- common_data_filtered %>%
  mutate(across(c(ART_Coverage, HIV_Incidence, AIDS_Mortality), ~ {
    num <- parse_number(., locale = locale(grouping_mark = " "))
    ifelse(str_detect(., "^<"), num / 2, num)
  }))

```

## Summary Table of Extreme Values
```{r}
summary_table <- data.frame(
  Indicator = c("Minimum ART Coverage", "Maximum ART Coverage", 
                "Minimum HIV Incidence", "Maximum HIV Incidence",
                "Minimum AIDS Mortality", "Maximum AIDS Mortality"),
  Value = c("1% (South Sudan, 2010)", 
            "96% (Rwanda, 2023)", 
            "100 (South Sudan, 2010)", 
            "7,700,000 (South Africa, 2023)", 
            "50 (Mauritania, 2010)", 
            "150,000 (South Africa, 2010)")
)

flextable(summary_table) %>%
  set_caption("Summary of Extreme Values for Key Indicators Across African Countries") %>%
  align(align = "left", part = "all")

```

## Export Summary Table to Word
```{r}
doc <- read_docx() %>%
  body_add_par("Summary Table of Extreme Values", style = "heading 1") %>%
  body_add_flextable(flextable(summary_table))

print(doc, target = "Summary_Table.docx")

```

## Correlation Tests
```{r}
# Shapiro-Wilk
shapiro.test(common_data_adjusted$ART_Coverage)
shapiro.test(common_data_adjusted$HIV_Incidence)
shapiro.test(common_data_adjusted$AIDS_Mortality)

# Spearman
cor.test(common_data_adjusted$ART_Coverage, common_data_adjusted$HIV_Incidence, method = "spearman")
cor.test(common_data_adjusted$ART_Coverage, common_data_adjusted$AIDS_Mortality, method = "spearman")
cor.test(common_data_adjusted$HIV_Incidence, common_data_adjusted$AIDS_Mortality, method = "spearman")

```
## Regressions
```{r}
# Basic Regression
summary(lm(AIDS_Mortality ~ ART_Coverage, data = common_data_adjusted))
summary(lm(AIDS_Mortality ~ HIV_Incidence, data = common_data_adjusted))
summary(lm(AIDS_Mortality ~ ART_Coverage + HIV_Incidence, data = common_data_adjusted))

# Panel Regression
panel_data <- pdata.frame(common_data_adjusted, index = c("Country", "Year"))
model_fe <- plm(AIDS_Mortality ~ ART_Coverage + HIV_Incidence, data = panel_data, model = "within")
summary(model_fe)

# Growth Curve Modeling
model_gcm1 <- lmer(AIDS_Mortality ~ ART_Coverage + Year + (1 | Country), data = common_data_adjusted)
summary(model_gcm1)

```
## Visualization – Trend Lines
```{r}
# ART
ggplot(common_data_adjusted, aes(x = Year, y = ART_Coverage)) +
  geom_line(color = "blue") +
  facet_wrap(~ Country) +
  labs(title = "ART Coverage by Country", y = "ART Coverage (%)") +
  theme_minimal()

# HIV Incidence
ggplot(common_data_adjusted, aes(x = Year, y = HIV_Incidence)) +
  geom_line(color = "red") +
  facet_wrap(~ Country) +
  labs(title = "HIV Incidence by Country", y = "HIV Incidence") +
  theme_minimal()

# AIDS Mortality
ggplot(common_data_adjusted, aes(x = Year, y = AIDS_Mortality)) +
  geom_line(color = "darkgreen") +
  facet_wrap(~ Country) +
  labs(title = "AIDS Mortality by Country", y = "AIDS Mortality") +
  theme_minimal()

```

## Map Creation (2023)
```{r}
# Fix country names for map match
data_2023 <- common_data_adjusted %>%
  filter(Year == 2023) %>%
  mutate(Country = case_when(
    Country == "South Sudan" ~ "S. Sudan",
    Country == "United Republic of Tanzania" ~ "Tanzania",
    Country == "Democratic Republic of the Congo" ~ "Dem. Rep. Congo",
    Country == "Eswatini" ~ "eSwatini",
    TRUE ~ Country
  ))

# Load shapefile
sf_use_s2(FALSE)
africa <- ne_countries(continent = "Africa", returnclass = "sf") |> st_make_valid()
map_data <- left_join(africa, data_2023, by = c("name" = "Country")) |> 
  st_transform(crs = 102022)

map_data$centroid <- st_centroid(map_data$geometry)
coords <- st_coordinates(map_data$centroid)
map_data$X <- coords[, 1]
map_data$Y <- coords[, 2]

```

## HIV Incidence Map
```{r}
ggplot(map_data) +
  geom_sf(aes(fill = HIV_Incidence), color = "grey50") +
  geom_text(aes(x = X, y = Y, label = name), size = 2) +
  scale_fill_viridis(option = "plasma", direction = -1, name = "HIV Incidence") +
  ggtitle("HIV Incidence in Africa (2023)") +
  theme_minimal()

```

## ART Coverage Map
```{r}
art_coverage_map <- ggplot(map_data) +
  geom_sf(aes(fill = ART_Coverage), color = "grey50", size = 0.3) +
  geom_text(aes(x = X, y = Y, label = name), size = 2, color = "black", fontface = "bold") +
  scale_fill_viridis(option = "cividis", direction = -1, na.value = "white",
                     name = "ART Coverage (%)", labels = scales::comma) +
  ggtitle("ART Coverage in Africa (2023)") +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank()
  )

# Display the map
print(art_coverage_map)

```
## AIDS-Related Mortality Map
```{r}
AIDS_Mortality_Map <- ggplot(map_data) +
  geom_sf(aes(fill = AIDS_Mortality), color = "grey50", size = 0.3) +
  geom_text(aes(x = X, y = Y, label = name), size = 2, color = "black", fontface = "bold") +
  scale_fill_viridis(option = "magma", direction = -1, na.value = "white",
                     name = "AIDS Mortality", labels = scales::comma) +
  ggtitle("AIDS-Related Mortality in Africa (2023)") +
  theme_minimal(base_size = 16) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank()
  )

# Display the map
print(AIDS_Mortality_Map)

```
## Save AIDS Mortality Map
```{r}
ggsave("AIDS_Mortality_Africa_2023.png", plot = AIDS_Mortality_Map, 
       path = "C:/Users/THIS PC/Downloads", width = 10, height = 7, dpi = 300)

```
## Save ART Coverage Map
```{r}
ggsave("ART_Coverage_Africa_2023.png", plot = art_coverage_map, 
       path = "C:/Users/THIS PC/Downloads", width = 10, height = 7, dpi = 300)

```

## Save HIV Incidence Map
```{r}
ggsave("HIV_Incidence_Africa_2023.png", plot = last_plot(), path = "C:/Users/THIS PC/Downloads", width = 10, height = 7, dpi = 300)

```

